Dynamical properties of the Benguela Current

### 2019-04-24 Write_up/Writing/dissertation.Rmd
update

---
title: "Dynamical properties of the Benguela Current as seen through a cold-spell algorithm."
author: "Amieroh Abrahams"
date: "13 January 2019"
output:
  pdf_document: default
  html_document: default
fontsize: 10pt
geometry: margin=1in
language: Australian
mainfont: Minion Pro
monofont: Courier
papersize: A4
sansfont: Myriad Pro
tables: yes
bibliography: ../bib_master.bib
---

# Chapter 1: Introduction

@@

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are shown to be forced by local wind stress fields off southwest Africa[@nelson; @fennel]. This EBUS streches along the south-western coast of Africa from southern Angola to Cape Agulhas [@cole]. Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms].  The upwelling period in the benguela north of 25°C  persist throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole]. Several research studies have shown extreme warming and cooling periods occurring in the northern and southern benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Research studies done by Walter (1937) also show anomously warm periods which may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming the equatorward wind stress over the northern region of the benguela system was stronger than usual; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occuring along the south coastline of the of Southern Africa are particularly generated by local winds.

Global warming is increasing at an extremely high rate. This will result in an increase in greenhouse gases which in turn will impact the ocean and terrestrial life. The process of upwelling is one such factor that is predicted to change with an increase in global warming. As prreviously mentioned, changes in temperature results in changes in pressure gradients, these pressure gradients control wind patterns which ultimately controls the intensity and duration of upwelling. To assess the changes within the upwelling events, various datasets would be used. These are satellite sea surface temperatures (broad scale), in situ collected seawater temperatures (local scale), wind data as well as satellite obtained chlorophyll-a data. Having assessed the eastern boundry upwelling system within a context that comprised of broad and local-scale features permitted us to assess how upwelling varied between sites located closer to the upwelling centres and furthur away from the upwelling centre. Specifically, we aim to assess whether or not it is possible to identify the upwelling signal throughout the various datasets and here we examine how far offshore the upwelling signal is present in the higher resolution data and how far offshore the SACTN signal extend. We hypothesized that (i) wind action may influence the intensity and duration of upwelling; (ii) Remotely sensed SST data will represent a similar event frequency and duration; (iii) Sites located closer to upwelling centres displays distinct variation in the intensity and duration of upwelling events compared to sites furthur away from the upwelling centeres.

@@

@@

>>>>

---
title: "Dynamical properties of the Benguela Current"
author: "Amieroh Abrahams"
date: "13 January 2019"
output:
  word_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    toc: yes
  html_document:
    fig_caption: yes
    fig_retina: 2
    fig_width: 7
    highlight: espresso
    theme: sandstone
    toc_float: no
  bookdown::html_document2:
    fig_caption: yes
    fig_retina: 2
    fig_width: 7
    highlight: espresso
    theme: sandstone
    toc_float: no
  pdf_document:
    toc: yes
    toc_depth: 4
    highlight: default
    keep_tex: yes
    latex_engine: xelatex
fontsize: 10pt
geometry: margin=1in
language: Australian
mainfont: Minion Pro
monofont: Courier
papersize: A4
sansfont: Myriad Pro
tables: yes
bibliography: ../bib_master.bib
---
# Chapter 1: Introduction

@@

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are shown to be forced by local wind stress fields off southwest Africa[@nelson; @fennel]. This EBUS streches along the south-western coast of Africa from southern Angola to Cape Agulhas [@cole]. Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms].  The upwelling period in the benguela north of 25°C  persist throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole]. Several research studies have shown extreme warming and cooling periods occurring in the northern and southern benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Research studies done by Walter (1937) also show anomously warm periods which may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming the equatorward wind stress over the northern region of the benguela system was stronger than usual; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occuring along the south coastline of the of Southern Africa are particularly generated by local winds. Event-scale upwelling has been demonstrated to have substantial impacts of individual on fisheries [@roy; @van]

Global warming is increasing at an extremely high rate. This will result in an increase in greenhouse gases which in turn will impact the ocean and terrestrial life. The process of upwelling is one such factor that is predicted to change with an increase in global warming. As previously mentioned, changes in temperature results in changes in pressure gradients, these pressure gradients control wind patterns which ultimately controls the intensity and duration of upwelling. To assess the changes within the upwelling events, various datasets would be used. These are satellite sea surface temperatures (broad scale), in situ collected seawater temperatures (local scale), wind data as well as satellite obtained chlorophyll-a data. Having assessed the eastern boundry upwelling system within a context that comprised of broad and local-scale features permitted us to assess how upwelling varied between sites located closer to the upwelling centres and furthur away from the upwelling centre. Specifically, we aim to assess whether or not it is possible to identify the upwelling signal throughout the various datasets and here we examine how far offshore the upwelling signal is present in the higher resolution data and how far offshore the SACTN signal extend. We hypothesized that (i) wind action may influence the intensity and duration of upwelling; (ii) Remotely sensed SST data will represent a similar event frequency and duration; (iii) Sites located closer to upwelling centres displays distinct variation in the intensity and duration of upwelling events compared to sites furthur away from the upwelling centeres.

@@

```{r , echo= FALSE, message=FALSE, warning=FALSE, fig.cap="Map representing the study region along the South African coastline. The black points represent the location of their *in situ* temperatures and approximations of the pixels used along the shore normal trasect from the satellite sea surface temperatures shown with black boxes", fig.height=8, fig.width=12, fig.pos="H"}

# knitr::include_graphics("../figures/map_fixed.pdf")
```

@@

## Results

### Wind patterns during upwelling events

```{r  echo= FALSE, message=FALSE, warning=FALSE, fig.cap="Wind rose diagrams representing the the most predominant wind direction for each of the sites for the years 2010 - 2014. Each spoke is divided by color into wind speed ranges. The radial length of each spoke around the circle is the amount of times that the wind comes from that direction", fig.height=8, fig.width=12, fig.pos="H"}

 knitr::include_graphics("../figures/wind.png")
```

## Local and broad scale upwelling
